---
title: "Missingness Report"
author: "Isaiah C. Mireles"
date: "2026-05-17"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---

# Read Data

```{r rd, include = F}
library(readr)
library(dplyr)
library(stringr)

blood <- 
  read_csv("~/Desktop/Material related to Dr. Ghahremani and Lavretsky/blood_data.csv")

diet <- 
  read_csv("~/Desktop/Material related to Dr. Ghahremani and Lavretsky/diet_data_update.csv")

blood <- blood[,-1]
diet <- diet[,-1]

relevant_dict <-
  read_csv("~/Desktop/Material related to Dr. Ghahremani and Lavretsky/relevant_dict.csv")
all <- read_csv("~/Desktop/Material related to Dr. Ghahremani and Lavretsky/Stats141XP_AABC_Release2_data.csv")

cl_nm <- all[1,]
all <- all[-1,]
colnames(all) <- cl_nm

library(readxl)
all_dict <- read_excel(
  "~/Desktop/Material related to Dr. Ghahremani and Lavretsky/Stats141XP_AABC_study_data_dictionary.xlsx"
)
```

# Order of Events

## event, study pairs

```{r}
all |> dplyr::distinct(event, study)
```

-   notice some peoples first visit ( `V1` ) may be in `HCA` or, `AABC` – same being true for `V2`

    -   suggesting people joined study at different time periods

```{r}
all <- all |>
  dplyr::mutate(
    event = factor(
      event,
      levels = c(
        "V1", "F1",
        "V2", "F2",
        "CR",
        "V3", "F3",
        "V4",
        "AF1"
      ),
      ordered = TRUE
    ),

    study = factor(
      study,
      levels = c("HCA", "AABC"),
      ordered = TRUE
    )
  )
```

```{r}
patient_paths <-
  all |>
  filter(age_open != "90 or older") |>
  mutate(age_open = as.numeric(age_open)) |> 
  mutate(
    patient = str_remove(id, "^[A-Z]+")
  ) |>
  distinct(patient, study, event, age_open) |>
  arrange(patient, study, event) |>
  group_by(patient) |>
  arrange(study, event) |> 
  summarise(
    n_studies = n_distinct(study),
    n_events = n(),

    years_passed = if_else(
      n() > 1,
      max(age_open, na.rm = TRUE) - min(age_open, na.rm = TRUE),
      NA_real_
    ),

    study_event_path = paste(
      paste(study, event, sep = ":"),
      collapse = " → "
    ),

    .groups = "drop"
  ) |>
  select(patient, years_passed, study_event_path, years_passed, everything()) |>
  arrange(desc(years_passed))
patient_paths
```

```{r}
plot(patient_paths$years_passed, patient_paths$n_events, xlab = "Yrs Passed", ylab = "n_events")
```

# Path Diagram

```{r}
patient_paths
```

```{r long_form, message = F}
library(tidyverse)

patient_steps <-
  patient_paths |>
  separate_rows(study_event_path, sep = " → ") |>
  separate(
    study_event_path,
    into = c("study", "event"),
    sep = ":"
  ) |>
  mutate(
    study = factor(
      study,
      levels = c("HCA", "AABC"),
      ordered = TRUE
    ),
    event = factor(
      event,
      levels = c(
        "V1", "F1",
        "V2", "F2",
        "CR",
        "V3", "F3",
        "V4",
        "AF1"
      ),
      ordered = TRUE
    )
  ) |>
  arrange(patient, study, event) |>
  group_by(patient) |>
  mutate(step = row_number()) |>
  ungroup()
```

```{r}
patient_steps <-
  patient_steps |>
  mutate(
    study_event = interaction(
      study,
      event,
      sep = ":",
      lex.order = TRUE
    )
  )
```

```{r}
patient_edges <-
  patient_steps |>
  arrange(patient, step) |>
  group_by(patient) |>
  mutate(
    next_study_event = lead(study_event),
    next_step = lead(step)
  ) |>
  filter(!is.na(next_study_event)) |>
  ungroup()
```

```{r}
transition_counts <-
  patient_edges |>
  count(
    step,
    next_step,
    study_event,
    next_study_event,
    name = "n_patients"
  )
```

```{r}
library(grid)

max_step <- max(patient_steps$step)

path_plot <-
  ggplot(transition_counts) +
  geom_segment(
    aes(
      x = step,
      xend = next_step,
      y = study_event,
      yend = next_study_event,
      color = n_patients,
      linewidth = n_patients
    ),
    arrow = arrow(length = unit(0.15, "cm")),
    alpha = 0.8
  ) +
  geom_point(
    data = patient_steps |> distinct(step, study_event),
    aes(x = step, y = study_event),
    size = 3
  ) +
  scale_x_continuous(
    breaks = seq(1, max_step, by = 1),
    limits = c(1, max_step)
  ) +
  scale_linewidth(range = c(0.5, 3)) +
  scale_color_gradient(
    low = "mistyrose",
    high = "red4"
  ) +
  labs(
    x = "Sequence Step",
    y = "Study:Event",
    color = "Number of Patients",
    linewidth = "Number of Patients",
    title = "Patient Study/Event Paths"
  ) +
  theme_minimal()

path_plot
```

```{r sv_file, include = F}
ggsave(
  filename = "patient_paths.png",
  plot = path_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

# Missingness Relevance

A great deal of data is missing ( `NA` ) – therefore our data is heavily **biased**. We would like to determine in *what way* its **biased.** Perhaps by adjusting for this bias we see strong associations for each variable to the cognitive markers in studying dimentia.

## Blood Data

```{r b_ms, message = F, warning=FALSE}
library(dplyr)
library(naniar)
blood |> 
  miss_var_summary() |>
  as.data.frame() |>
  select(variable, pct_miss) 
```

-   direct_ldl is entirely missing, we may discard this feature

-   Many repeated values

## Diet Data

```{r}
diet |> 
  miss_var_summary() |>
  as.data.frame() |>
  select(variable, pct_miss) 
```

[interpretation]{.underline} :

-   Notice the only features without missingness

-   Notice the large amount of missingness is the

    |                               |
    |:------------------------------|
    | Participant info/Demographics |

    -   **`data_category`**

-   Notice that the missingness appears to repeat – showing *runs* of the same `pct_miss` for both features

## Ct and Graph Repeated Missingness

## Diet Data

```{r}
ct <-
  diet |> 
  miss_var_summary() |> 
  mutate(
    ct = ifelse(pct_miss == 0, 0, 1)
    ) |> count(ct) 
ct
```

[Interpretation]{.underline} :

-   There are only 4 non-missing values every other value is just `84.6` percent missing – which is a **HUGE** amt.

## Blood Data

```{r}
ct <- 
  blood |> 
  miss_var_summary() 

library(ggplot2)
ct |> 
  ggplot(aes(pct_miss)) + 
  geom_histogram(binwidth = 10) 
```

[interpretation]{.underline}

-   this one is slightly more varied

-   0, about 60, 90 100% percent missing – with 60, 90 happening quite frequently

```{r}
ct <- ct |>
  mutate(
    char = format(pct_miss)
  )

ct <- as.vector(ct$pct_miss)/100
ct <- ct |> round(1) 
ct <- ct |> as.character()
barplot(table(ct))
```

-   We can see that the repeated values are near 0 (id features), 60%, 90% or 100%

# Build Features

## `patient` **study & events ct**

[Purpose]{.underline} : identify how often a `patients` appear in our data.

For example :

-   Are they in both studies?

-   How many unique events?

### Helper Func : Create `unique_id`, `Patient` Variable

```{r}
library(dplyr)

diet <-
  diet |>
  mutate(
    patient = str_remove(id, "^[A-Z]+") 
  ) |> select(patient, everything())

blood <-
  blood |>
  mutate(
    patient = str_remove(id, "^[A-Z]+") 
  ) |> select(patient, everything())
```

# `patient` specific missing ct

[Purpose]{.underline} : identify how often `patients` appear in our data to better understand missing values.

## Diet

```{r}
diet_patients <-
  diet |>
  group_by(patient) |>
  summarize(
    n_rows = n(),
    total_missing =
      sum(across(everything(), ~ sum(is.na(.x)))),
    .groups = "drop"
  ) |>
  mutate(
    pct =
      round(
        total_missing /
          (n_rows * (ncol(diet) - 1)),
        2
      )
  ) |>
  arrange(desc(pct))

diet_patients <-
  diet_patients |>  mutate(
    p_rank = paste0("patient", row_number())
  ) |> select(p_rank, everything(), patient)
```

```{r}
plot(diet_patients$n_rows, diet_patients$total_missing)
```

## blood

```{r}
blood_patients <-
  blood |>
  group_by(patient) |>
  summarize(
    n_rows = n(),
    total_missing =
      sum(across(everything(), ~ sum(is.na(.x)))),
    total_cells =
      n_rows * ncol(pick(everything())),
    pct =
      round(total_missing / total_cells, 2),
    .groups = "drop"
  ) |>
  arrange(desc(pct))

blood_patients <-
  blood_patients |>  mutate(
    p_rank = paste0("patient", row_number())
  ) |> select(p_rank, everything(), patient)
```

```{r}
plot(blood_patients$n_rows,blood_patients$total_missing)
```

# Correlations of `Cognitive` Features – ALL

Note :

-   we have repeated observations, so we are taking people who had more visits more into account

```{r str_splt, include = F}
#grep("dict", ls(), value=T)
all_dict |> 
  filter(data_category=="Cognitive") |> 
  select("variable name - description") 
```

-   Relevant factor analysis Features\^

    -   `Memory_Tr35_60y`,

        -   Cognition Factor Analysis: Memory factor estimated in Union V1 35-60 yo subjects and applied to all subjects all visits

    -   `FluidIQ_Tr35_60y`,

        -   Cognition Factor Analysis: Fluid IQ factor estimated in Union V1 35-60 yo subjects and applied to all subjects all visits

    -   `CrystIQ_Tr35_60y`

        -   Cognition Factor Analysis: Crystallized IQ factor estimated in Union V1 35-60 yo subjects and applied to all subjects all visits

```{r}
Memory <- all$Memory_Tr35_60y |> as.numeric()
fluidIQ <- all$FluidIQ_Tr35_60y |> as.numeric()
CrystIQ <- all$CrystIQ_Tr35_60y |> as.numeric()
```

## `Memory`, `fluidIQ` dot-plt

```{r}
# Memory ~ fluidIQ
Memory_fluidIQ_mdl <- lm(Memory ~ fluidIQ)

plot(Memory, fluidIQ)
abline(Memory_fluidIQ_mdl, col = "red")
```

## `Memory`, `CrystIQ` dot-plt

```{r}
# Memory ~ CrystIQ
Memory_CrystIQ_mdl <- lm(Memory ~ CrystIQ)

plot(Memory, CrystIQ)
abline(Memory_CrystIQ_mdl, col = "red")
```

## `fluidIQ`, `CrystIQ` dot-plt

```{r}
# fluidIQ ~ CrystIQ
fluidIQ_CrystIQ_mdl <- lm(fluidIQ ~ CrystIQ)

plot(fluidIQ, CrystIQ)
abline(fluidIQ_CrystIQ_mdl, col = "red")
```

```{r}
diet_num <- diet |> select(where(is.numeric))
blood_num <- blood |> select(where(is.numeric))
```

```{r}
diet_num <- cbind(Memory, fluidIQ, CrystIQ, diet_num)
blood_num <- cbind(Memory, fluidIQ, CrystIQ, blood_num)
```

# Diet Correlations w/ `Memory`

```{r}
mem_cor <-
  cor(diet_num$Memory, diet_num[4:length(diet_num)],
      use = "complete.obs")


mem_cor <-
  mem_cor |> 
  t() |> as.data.frame()
```

```{r}
mem_cor <- 
  mem_cor |> 
  rename(Memory = V1) |> 
  round(3) |>
  select(Memory)

mem_cor 
```

# Diet Correlations w/ `fluidIQ`

```{r}
fluid_cor <-
  cor(diet_num$fluidIQ, diet_num[4:length(diet_num)],
      use = "complete.obs")


fluid_cor <-
  fluid_cor |> 
  t() |> as.data.frame()

fluid_cor |> 
  rename(fluidIQ = V1) |> 
  round(3) |>
  select(fluidIQ)
```

# Diet Correlations w/ `CrystIQ`

```{r}
cryst_cor <-
  cor(diet_num$CrystIQ, diet_num[4:length(diet_num)],
      use = "complete.obs")


cryst_cor <-
  cryst_cor |> 
  t() |> as.data.frame()

cryst_cor <-
  cryst_cor |> 
  rename(CrystIQ = V1) |> 
  round(3) |> select(CrystIQ) 


cryst_cor
```

[Overall interpretation]{.underline}

-   Extremely weak correlations for all variables in Diet Data

    -   Diet Data doesnt appear strongly related

[Possible Causes]{.underline}

-   underlying relationship isnt linear

-   Missing values may be biasing variables

```{r}
diet_cor_lst <-
  list(
  cryst_cor=cryst_cor,
  fluid_cor=fluid_cor,
  mem_cor=mem_cor
)
```

# Patient level analysis

```{r}
blood |>
  group_by(patient) |> 
  select(patient, where(is.numeric)) |> 
  summarize(
    across(everything(), ~ mean(.x, na.rm = TRUE))
  )
```

# Patient Correlations of `Cognitive` Features

-   here we generate a report of correlations for each individual patient

-   Suppose we just average out their values across all `Study` and `event`

```{r}
diet_patients
```

# Multivariate Normal – Cognitive Features

```{r}
cognitive_df <- 
  all |> 
  select(id_event, id, event, study,
         Memory_Tr35_60y, 
         FluidIQ_Tr35_60y, 
         CrystIQ_Tr35_60y) |>
  mutate(
    patient = str_remove(id, "^[A-Z]+") 
  ) |> select(patient, everything()) |>
  mutate(
    across(
      -c(patient, id_event, id, event, study),
      as.numeric
    )
  )
```

```{r}
par(mfrow = c(1, 3))

hist(
  cognitive_df$Memory_Tr35_60y,
  main = "Memory",
  xlab = ""
)

hist(
  cognitive_df$FluidIQ_Tr35_60y,
  main = "FluidIQ",
  xlab = ""
)

hist(
  cognitive_df$CrystIQ_Tr35_60y,
  main = "CrystIQ",
  xlab = ""
)
```

-   notice each is normally distributed

```{r}

cognitive_df |>
  select(where(is.numeric)) |>
  summarize(
    mean_Memory = mean(Memory_Tr35_60y, na.rm = TRUE), 
    sd_Memory = sd(Memory_Tr35_60y, na.rm = TRUE), 
    mean_FluidIQ_y = mean(FluidIQ_Tr35_60y, na.rm = TRUE),
    sd_FluidIQ = sd(FluidIQ_Tr35_60y, na.rm = TRUE), 
    mean_CrystIQ = mean(CrystIQ_Tr35_60y, na.rm = TRUE),
    sd_CrystIQ = sd(CrystIQ_Tr35_60y, na.rm = TRUE)
  )
```
